### 2014-05-30 patchy-selection-paper.tex
edited the haplotype sharing section

@@

If a patch adapts through new mutation or a rare migrant lineage, the
genomic region surrounding the selected site will hitchhike along with
it \citep{maynardsmith1974hitchhiking},
so that at least initially, all adapted individuals within a patch
share a fairly long stretch of haplotype. 
Pairs of individuals within a patch will share a haplotype of mean genetic length of
$\approx \log(\rho A s_p)/s_p$ around the selected site, 
as long as the patch is reasonably well mixed by dispersal
\citep[otherwise see][]{barton2013genetic}.
This association gets slowly whittled down by recombination during interbreeding at the edge of the patch;
but there will always be longer LD nearby to the selected site. 
\plr{Could work out on what time scale it gets whittled down and the shared length at stationarity.}

When an already adapted patch colonizes another through migration,
then the newly colonized patch will inherit a long piece of haplotype around the selected site from the originating patch.
The genetic length of this haplotype is roughly inversely proportional to the
number of generations the allele spends ``in transit'', 
because each reproductive event in locations where the allele is at low frequency recombines haplotypes of the $b$ allele closer to the adaptive allele.
(In this case, the haplotype is literally hitchhiking across geography!)

@@

As a general rule of thumb, equation~\eqref{eqn:mean_tau} means that families who successfully establish
move at speed $\sqrt{s_m} \sigma$ towards the new patch 
-- if $s_m$, the strength of the selection against them, is smaller, the need to move quickly is less imperative.
Then, equation~\eqref{eqn:haplotype_length} means that the haplotype length
is roughly the length one would expect given the mean transit time.
Note that we have become used to seeing $\sigma$ divided by $\sqrt{s_m}$, rather than multiplied,
the $\sigma/\sqrt{s_m}$ is a length scale, and to convert it to a time this is divided by $1/s_m$.

%%%
\subsubsection{Limiting distribution of the transit time and hitchhiking haplotype}

@@

\plr{let's leave two dimensions for the revisions. or, as an exercise? =)}

@@

To really make use of the expressions in the previous section,
we need to know the length of the haplotype shared by individuals \emph{within} a patch.
A detailed treatment of this is well beyond the scope of this paper,
but we can get a good idea of the scale through heuristic arguments.
To do this, we compare the rate of recombination between B and b haplotypes
to the rate of coalescence of B haplotypes within a patch,
which we take to be $1/(A\rho)$ 
(bypassing the well-known difficulties to spatial coalescence \citep{felsenstein1978pain,bartonetheridgedepaulis}).

Consider the lineage of a single B allele, moving back through time,
in a patch at mutation-selection balance, 
i.e.\ with the local proportion of B alleles near $x$ equal to $p(x)$,
and that the lineage stays close enough to the patch 
that fluctuations about $p(x)$ are small (unlike in previous sections).
If we suppose that type B individuals at location $y$ reproduce at rate $1+s(y)$,
and the chance their offspring move to $x$ is $m(y,x)$ (with $\sum_x m(y,x)=1$),
then the total rate of influx of B alleles from $y$ into $x$ is $\rho p(y) (1+s(y)) m(y,x)$.
As argued by \citet{hudson}, the rate at which a B lineage moves from $x$ to $y$
is this influx divided by the local abundance of $B$ alleles,
or $\frac{ p(y) (1+s(y)) m(y,x) }{ p(x) }$.
After the same large-population, continuous-space limit that underlies the rest of this paper,
this motion should converge to a diffusion process \citep[as assumed by][]{hallatschek},
notwithstanding some technical difficulties (e.g.\ discontinuity of $s(x)$).
In It\^o differential notation, 
the location $X_t$ of a B lineage $t$ generations ago satisfies
\begin{align}
    d X_t &= (1+s(X_t)) \frac{d}{d\,x} \log( p(X_t) (1+s(X_t)) ) dt + \sigma \sqrt{1+s(X_t)} dB_t,\\
    % &=  ( \log(p(X_t))' (1+s(X_t)) + s'(X_t) ) dt + \sqrt{1+s(X_t)} dB_t,\\
\end{align}
where $s(x)=s_p$ if $x$ is in the patch and $s(x)=s_m$ otherwise, 
and $B_t$ is a standard Brownian motion.
This is a diffusion that, when at $x$, 
is pulled toward regions with more B alleles
by the drift term $(1+s(x))\frac{d}{d\,x} \log( p(x) (1+s(x)) )$,
and moves at nearly constant speed (since $\sqrt{1+s(x)}$ is close to 1).
Furthermore, it is easy to verify that the stationary distribution of this process is
\begin{align}
  \pi(x) = \frac{ p(x)^2 (1+s(x)) }{ \int_y p(y)^2 (1+s(y)) dy } .
\end{align}

A lineage at a locus at genetic distance $r$ that begins linked to a B allele
moves as a B lineage up until the most recent time
it recombined from a b background to a B background,
which occurs when at $x$ with rate $r (1-p(x)) (1+s(x)/2)$.
The Feynman-Kac formula could then be used to write down differential equations
solved by various moments of the haplotype length as a function of the sampling location of the lineage.

Or, more simply, if recombination is on a slower time scale than mixing within the patch,
then the time until the last recombination is the integral of $r (1-p(x)) (1+s(x)/2)$ 
against the stationary distribution.
Within the patch, $p(x) \approx 1$,
so recombination between backgrounds is unlikely to occur,
and so the rate of recombination depends to a first approximation 
on the proportion of the patch in which $p(x)$ is intermediate.
This region where $p(x)$ is intermediate has width of order $\sigma/\sqrt{s}$, where $s = \min(s_p,s_m)$.
As we increase the size of a patch, the area on which $p(x)$ is intermediate scales with the perimeter of the patch.
Therefore, the proportion of the time a lineage at stationarity is found in the intermediate regions
is roughly $2(\sigma/\sqrt{s})A^{-1/d}$,
and so the rate of recombination of a linked lineage to the b background
is, taking $p(x) \approx 1/2$ in this region, roughly $r (\sigma/\sqrt{s})A^{-1/d}$,
This will be equal to the coalescence rate if
$1/(A\rho) = r (\sigma/\sqrt{s}) A^{-1/d}$, 
and hence the length of linkage shared within the patch is on the order of
\begin{align}
  \frac{1}{r} = \frac{ \sigma \rho A^{(d-1)/d} }{\sqrt{s}} ,
\end{align}
which does not increase with $A$ in one dimension,
but increases with $\sqrt{A}$ in two dimensions.

@@

>>>>

\begin{figure}[ht]
  \begin{center}
    \includegraphics{sim-transit.pdf}
  \end{center}
  \caption{
  BLAH BLAH WIGGLY LINE
  \label{fig:lineagesmotion}
  }
\end{figure}

@@

If a patch adapts through new mutation or a rare migrant lineage, the
genomic region surrounding the selected site will hitchhike along with
it \citep{maynardsmith1974hitchhiking},
so that at least initially, all adapted individuals within a patch
share a fairly long stretch of haplotype. 
Pairs of individuals within a patch will share a haplotype of mean genetic length of
$\approx \log(\rho A s_p)/s_p$ around the selected site, 
as long as the patch is reasonably well mixed by dispersal
\citep[otherwise see][]{barton2013genetic}.
This association gets slowly whittled down by recombination during interbreeding at the edge of the patch;
but there will always be longer LD nearby to the selected site.

When an already adapted patch colonizes another through migration,
then the newly colonized patch will inherit a long piece of haplotype around the selected site from the originating patch.
The genetic length of this haplotype is roughly inversely proportional to the
number of generations the allele spends ``in transit'', 
because each reproductive event in locations where the allele is at
low frequency recombines haplotypes of the $b$ allele closer to the
adaptive allele. In this case, the haplotype is literally hitchhiking across geography!
\gc{In Figure \ref{fig:lineagesmotion} we show the lineage that founds a
mutant population on a second patch. The entire time that the lineage
is outside the dark red zone (where the B allele is common) the
haplotype that accompanies it is being broken down. After the lineage
establishes on the patch the rate of decay of the haplotype is
significant slowed as it is again surrounded by other B
gneotypes with the similar haplotypic backgrounds. We first discuss
the transit of the lineage, and its accompanying haplotype, and return
to this point about the whittling down of haplotypes within
estalbished patches in the next section.}

@@

As a general rule of thumb, equation~\eqref{eqn:mean_tau} means that families who successfully establish
move at speed $\sqrt{s_m} \sigma$ towards the new patch 
-- if $s_m$, the strength of the selection against them, is smaller, the need to move quickly is less imperative.
Then, equation~\eqref{eqn:haplotype_length} means that the haplotype length
is roughly the length one would expect given the mean transit time.
Note that we have become used to seeing $\sigma$ divided by $\sqrt{s_m}$, rather than multiplied,
the $\sigma/\sqrt{s_m}$ is a length scale, and to convert it to a time this is divided by $1/s_m$.
\gc{In the Appendix Section \ref{apx:limiting_dist_of_HH_haplotype} we
  give some expressions for the limiting distribution of
  the length of shared haplotype.}

\gc{Graham pasted this to the appendix as it felt like too much information to drop on
  people}
%%%

@@

@@

The other key issue in thinking about the population genetic signal of
adaptation to spatial patches is the process by which haplotypes are
whittled down within patches, which plays into a number of aspects of
haplotype sharing. Thee initial haplotype of the sweep within a patch will
diserpate over time as recombination brokes up the haplotype; likewise the haplotype that is shared 
between patches that have
adapted through a shared migrant allele
(eqn. \ref{eqn:haplotype_length}) will also breakdown. 
We first develop a simple approximation to the rate of
recombination controling the breakdown of haplotypes within patches. 
If we wait long enough after the initial sweep,
either due to migration or mutation, the initial pattern will have
decayed due to this process of recombination. However, within patches
we should expect to find a larger extent of haplotype sharing close to
the selected locus, than in surrounding geographic regions as
selection against migrant lineage slows the diserpation of
genetic drift accumulated within the patch \citep{bartonpaper,charlesworths}.

We refer the reader again to Figure \ref{fig:lineagesmotion}. When a lineage in
established patch is in the dark red area, where the B allele is common, it is very unlikely to
recombine off the B background. However, when it wanders out of the
patch into the paler red and white areas when it recombines it will
usually be on to the b allele background.

A detailed treatment of this is well beyond the scope of this paper,
but we can get a good idea of the genomic scale of these effects through heuristic arguments.
To understand the probability of a recombination moving our lineage
off of the B background we need to understand what fraction of the
time our lineage spends in geographic regions where $b$ in common. 
Consider the lineage of a single B allele, moving back through time,
in a patch at migration-selection balance, 
i.e.\ with the local proportion of B alleles near $x$ equal to $p(x)$,
and that the lineage stays close enough to the patch 
that fluctuations about $p(x)$ are small. (Note that this assumption differs from
previous sections, but here our focus is on a typical lineage in the
patch rather than the rare lineages that make it far from the patch).
If we suppose that type B individuals at location $y$ reproduce at rate $1+s(y)$,
and the chance their offspring move to $x$ is $m(y,x)$ (with $\sum_x m(y,x)=1$),
then the total rate of influx of B alleles from $y$ into $x$ is $\rho p(y) (1+s(y)) m(y,x)$.
As argued by \citet{hudson}, the rate at which a B lineage moves from $x$ to $y$
is this influx divided by the local abundance of $B$ alleles,
or $\frac{ p(y) (1+s(y)) m(y,x) }{ p(x) }$.
After the same large-population, continuous-space limit that underlies the rest of this paper,
this motion should converge to a diffusion process \citep[as assumed by][]{hallatschek},
notwithstanding some technical difficulties (e.g.\ discontinuity of $s(x)$).
In It\^o differential notation, 
the location $X_t$ of a B lineage $t$ generations ago satisfies
\begin{align}
    d X_t &= (1+s(X_t)) \frac{d}{d\,x} \log( p(X_t) (1+s(X_t)) ) dt + \sigma \sqrt{1+s(X_t)} dB_t,\\
    % &=  ( \log(p(X_t))' (1+s(X_t)) + s'(X_t) ) dt + \sqrt{1+s(X_t)} dB_t,\\
\end{align}
where $s(x)=s_p$ if $x$ is in the patch and $s(x)=s_m$ otherwise, 
and $B_t$ is a standard Brownian motion.
This is a diffusion describing the movement of our lineage backward in
timethat, when at $x$, 
is pulled toward regions with more B alleles
by the drift term $(1+s(x))\frac{d}{d\,x} \log( p(x) (1+s(x)) )$,
and moves at nearly constant speed (since $\sqrt{1+s(x)}$ is close to 1).
Furthermore, it is easy to verify that the stationary distribution of this process is
\begin{align}
  \pi(x) = \frac{ p(x)^2 (1+s(x)) }{ \int_y p(y)^2 (1+s(y)) dy } . \label{eqn:lineagestatdist}
\end{align}
A lineage at a locus at genetic distance $r$ that begins linked to a B allele
moves as a B lineage up until the most recent time
it recombined from a b background to a B background,
which occurs when at $x$ with rate $r (1-p(x)) (1+s(x)/2)$.
The Feynman-Kac formula could then be used to write down differential equations
solved by various moments of the haplotype length as a function of the sampling location of the lineage.

As our goal here to describe the rough genomic length scale, 
based on the above, we view 
the lineage is simply bouncing around in the patch backward in time and we propose a
simple heurstic to describe the probability of recombination. 
If recombination is on a slower time scale than mixing within the patch,
then the time until the last recombination is the integral of $r (1-p(x)) (1+s(x)/2)$ 
against the stationary distribution (eqn. \ref{eqn:lineagestatdist}).
Within the patch, $p(x) \approx 1$,
so recombination off the $B$ background is unlikely to occur,
and so the rate of recombination depends to a first approximation 
on the proportion of the patch in which $p(x)$ is intermediate,
i.e. were both $B$ and $b$ are present.
This region where $p(x)$ is intermediate has width of order $\sigma/\sqrt{s}$, where $s = \min(s_p,s_m)$.
As we increase the size of a patch, the area on which $p(x)$ is intermediate scales with the perimeter of the patch.
Therefore, the proportion of the time a lineage at stationarity is found in the intermediate regions
is roughly $2(\sigma/\sqrt{s})A^{-1/d}$, and in these regions $p(x)
\approx 1/2$. Under these approximations the per generation rate of
recombination of a linked lineage to the b background is, taking in this region, roughly 
\begin{equation}
r (\sigma/\sqrt{s})A^{-1/d} \label{eqn:rate-rec-in-patch}
\end{equation}
We can use this to first of all think about the length of haplotype
associated with the $B$ allele after a given length of time $\tau$,
this could the the time since the sweep within a patch due to a
mutation or migration of a beneficial allele. If the recombination
rate per base is $r_{BP}$ then the length of haplotype associated with
the $B$ allele is
\begin{equation}
\frac{\sqrt{s}}{\sigma A^{-1/d}}\frac{1}{r_{BP}}
\end{equation} 
note that the length of haplotype decays at a rate proportional to the
cline width $\sqrt{s} / \sigma$, as the narrower the cline the less
opportunity for recombination to move our lineage off the $B$
background.

Finally we can consider the expected length of haplotype shared within
a patch. To do this, we compare the rate of recombination between B
and b haplotypes (eqn. \ref{eqn:rate-rec-in-patch})
to the rate of coalescence of B haplotypes within a patch. We will
assume that mixing within the patch is fast compared to the rate of
drift, so that the rate of coalescence of B alleles within a patch is 
$1/(A\rho)$ 
(this bypasses the well-known difficulties to spatial coalescence \citep{felsenstein1978pain,bartonetheridgedepaulis}).
The rate of recombination and coalescence will be on the same order if
$\xi /(A\rho) \approx r (\sigma/\sqrt{s}) A^{-1/d}$, 
and hence the genetic length of haplotype shared between pairs of
individuals within the patches is on the order of
\begin{align}
  r = \frac{\sqrt{s} \xi}{\sigma \rho A^{(d-1)/d}} 
\end{align}
which does not increase with $A$ in one dimension,
but decreases with $\sqrt{A}$ in two dimensions .

@@

\section{Limiting distribution of the transit time and hitchhiking haplotype}
\label{apx:limiting_dist_of_HH_haplotype}
